Combining living and fossil taxa into phylogenies: the missing data issue

### 2015-01-27 Manuscript/CoverLetter.tex
Cover letter update
Updated cite: Jackson -> Quental (more recent)
Fixed capital letters for Tip-Dating and abstract

In recent years there has been growing interest in building phylogenies that contain both living and fossil taxa (e.g. Fritz et al. 2013 TREE; Heath et al. 2014 PNAS). Such phylogenies could revolutionize the way we think about macroevolutionary patterns and processes, and provide a more complete understanding of trends in biodiversity through time. Unfortunately building such trees has proved technically difficult.

@@

Our results suggest that increasing the number of taxa with morphological data and the overall number of morphological characters will greatly improve the quality of Total Evidence tree topologies. This has major implications for clades where detailed morphological data collection from living species is rare. Additionally, we recommend using a Bayesian majority consensus tree when fixing tree topology for any additional analyses such as Tip-Dating (Ronquist et al. 2012 Syst Biol).

We feel this paper will be of particular interest to the readers of Systematic Biology as many papers investigating methods for combining living and fossil taxa and the effects of missing data on such analyses (e.g. D\'{a}valos et al. 2014; Pattinson et al. in press, Pyron 2011, Ronquist et al. 2012) % Any more? Without going on to a second page?
, have been recently published in the journal.

>>>>

In recent years there has been growing interest in building phylogenies that contain both living and fossil taxa (e.g. Quental and Marshall 2006 TREE; Fritz et al. 2013 TREE; Heath et al. 2014 PNAS). Such phylogenies could revolutionize the way we think about macroevolutionary patterns and processes, and provide a more complete understanding of trends in biodiversity through time. Unfortunately building such trees has proved technically difficult.

@@

Our results suggest that increasing the number of taxa with morphological data and the overall number of morphological characters will greatly improve the quality of Total Evidence tree topologies. This has major implications for clades where detailed morphological data collection from living species is rare. Additionally, we recommend using a Bayesian majority consensus tree when fixing tree topology for any additional analyses such as tip-dating (Ronquist et al. 2012 Syst Biol).

We feel this paper will be of particular interest to the readers of Systematic Biology as many papers investigating methods for combining living and fossil taxa and the effects of missing data on such analyses (e.g. D\'{a}valos et al. 2014; Pattinson et al. in press, Sylvestro et al. 2014, Pyron 2011, Ronquist et al. 2012), have been recently published in the journal.

### 2015-01-27 Manuscript/TEM_draft.tex
Fixed capital letters for Tip-Dating and abstract

\newpage
\begin{abstract}
To fully understand macroevolutionary patterns and processes, we need to include both extant and extinct species in our models. This requires phylogenetic trees with both living and fossil taxa at the tips. One way to infer such phylogenies is the Total Evidence approach which uses molecular data from living taxa and morphological data from living and fossil taxa. Although the Total Evidence approach is very promising, it requires a lot of data and is therefore likely to suffer from missing data issues that may affect its ability to infer correct phylogenies.

Here we use simulations to assess the effects of missing data on tree topologies inferred from Total Evidence matrices. We investigate three major factors that directly affect the completeness of the morphological part of the matrix: the proportion of living taxa with no morphological data, the amount of missing data in the fossil record, and the overall number of morphological characters in the matrix. We infer phylogenies from complete matrices and from matrices with various amounts of missing data, and then compare missing data topologies to the "best" tree topology inferred using the complete matrix.

We find that the number of living taxa with morphological characters and the number of morphological characters in the matrix, are more important than the amount of missing data in the fossil record for recovering the "best" tree topology. Therefore, we suggest that sampling effort should be focused on morphological data collection for living species to increase the accuracy of topological inference in a Total Evidence framework. Additionally, we find that Bayesian methods consistently outperform other tree inference methods. We therefore recommend using Bayesian consensus trees to fix the tree topology prior to further analyses such as Tip-Dating.

@@

A final approach, known as the Total Evidence method, uses matrices containing molecular data from living taxa and morphological data from both living and fossil taxa \citep{eernissetaxonomic1993}. This approach treats every taxon as a tip in the phylogeny, uses the occurrence age of the fossils to time calibrate the phylogeny (known as Tip-Dating; \citealt{ronquista2012}), and allows the use of probabilistic methods for estimating phylogenetic uncertainty \citep{ronquista2012}. Total Evidence methods have been successfully applied to empirical data \citep[e.g.][]{pyrondivergence2011,ronquista2012,schragocombining2013,slaterphylogenetic2013,beckancient2014}, and are becoming an increasingly popular way of adding fossil taxa to phylogenies. However, although the Total Evidence approach seems very promising, there is one big drawback in using this approach: it requires a lot of data that can be difficult (or impossible) to collect. The morphological data for living taxa are rarely collected when molecular data are available (e.g. \citealt{O'Leary08022013} \textit{vs.} \citealt{meredithimpacts2011}), and for fossil taxa, data can only be collected from features preserved in the fossil record (for example, in vertebrates, the hardest parts of the skeleton are more often preserved than soft parts; \citealt{sansomfossilization2013}). Therefore Total Evidence matrices are likely to contain a lot of missing data that may affect the method's ability to infer correct topologies, branch lengths and support values \citep{salamin2003}.

@@

Another practical implication of our results regards the tree inference methods. Because the Bayesian consensus trees consistently recovered topologies closer to the "best" tree topology than the Maximum Likelihood trees, we advise using Bayesian consensus trees to fix the topology where necessary for further steps in tree inference, for example in Tip-Dating \citep{ronquista2012,BEASTmaster}. Note that, although the Bayesian posterior tree distribution performed poorly in recovering the "best" tree topology, this is due to difficulties comparing distributions of trees (see discussion above). Thus, we do not suggest discarding the Bayesian posterior tree distribution once the topology has been fixed, particularly because these trees will be invaluable for phylogenetic comparative analyses \citep[e.g.][]{jetzthe2012}.

@@

Missing data in Total Evidence matrices is not a problem for recovering the "best" tree topology as long as enough living and fossil taxa in the matrix have data for overlapping morphological characters. When missing data increases in any of our missing data parameters ($M_{L}$, $M_{F}$ or $M_{C}$), it reduces support for the placement of fossil taxa and increases the displacement of wildcard taxa. Therefore we advise filling as many gaps in Total Evidence matrices as possible. Because this is difficult, if not impossible, for fossil data, we recommend coding as many morphological characters, for as many living taxa, as possible. Additionally, the topology of the Bayesian consensus trees, regardless the amount of missing data, were always closer to the "best" tree topology than the Maximum Likelihood trees. Therefore, we advise using Bayesian consensus tree topologies if topology must be fixed in a Total Evidence analysis, for example when using the Tip-Dating method. The results of our analyses are encouraging and show that it is possible to combine both neontological and palaeontological data in the same phylogeny despite issues of missing data. Hopefully, using these approaches will greatly improve our understanding of macroevolutionary processes through both time and space.

>>>>

\newpage
\begin{abstract}
To fully understand macroevolutionary patterns and processes, we need to include both extant and extinct species in our models. This requires phylogenetic trees with both living and fossil taxa at the tips. One way to infer such phylogenies is the Total Evidence approach which uses molecular data from living taxa and morphological data from living and fossil taxa.
%V1
Although the Total Evidence approach is very promising, it requires many data and is therefore likely to suffer from missing data issues that may affect its ability to infer correct phylogenies.
%V2
Although the Total Evidence approach is very promising, morphological characters are often limited and missing, therefore this method is likely to suffer from missing data issues that may affect its ability to infer correct phylogenies.

Here we use simulations to assess the effects of missing data on tree topologies inferred from Total Evidence matrices. We investigate three major factors that directly affect the completeness 
%How does the overall number of characters affect completeness of the matrix?
and the size
of the morphological part of the matrix: the proportion of living taxa with no morphological data, the amount of missing data in the fossil record, and the overall number of morphological characters in the matrix. We infer phylogenies from complete matrices and from matrices with various amounts of missing data, and then compare missing data topologies to the "best" tree topology inferred using the complete matrix.

We find that the number of living taxa with morphological characters and the overall number of morphological characters in the matrix, are more important than the amount of missing data in the fossil record for recovering the "best" tree topology
%Nick's comment
for a given number of missing characters.
Therefore, we suggest that sampling effort should be focused on morphological data collection for living species to increase the accuracy of topological inference in a Total Evidence framework. Additionally, we find that Bayesian methods consistently outperform other tree inference methods. We therefore recommend using Bayesian consensus trees to fix the tree topology prior to further analyses such as tip-dating.
%Hummm Nick's comment again - That's a point to make in the discussion maybe. But let's wait for Gavin comments...
%it's possible that adding dating information etc. in tip-dating makes Bayesian dated total-evidence analyses (ie tip-dating) better than non-dated Bayesian total-evidence analyses (typical total-evidence approach), because more information is being used, and because dates give some information about phylogenetic position.  In any case, the study wasn't about tip-dating

@@

A final approach, known as the Total Evidence method, uses matrices containing molecular data from living taxa and morphological data from both living and fossil taxa \citep{eernissetaxonomic1993}. This approach treats every taxon as a tip in the phylogeny, uses the occurrence age of the fossils to time calibrate the phylogeny (known as tip-dating; \citealt{ronquista2012}), and allows the use of probabilistic methods for estimating phylogenetic uncertainty \citep{ronquista2012}. Total Evidence methods have been successfully applied to empirical data \citep[e.g.][]{pyrondivergence2011,ronquista2012,schragocombining2013,slaterphylogenetic2013,beckancient2014}, and are becoming an increasingly popular way of adding fossil taxa to phylogenies. However, although the Total Evidence approach seems very promising, there is one big drawback in using this approach: it requires a lot of data that can be difficult (or impossible) to collect. The morphological data for living taxa are rarely collected when molecular data are available (e.g. \citealt{O'Leary08022013} \textit{vs.} \citealt{meredithimpacts2011}), and for fossil taxa, data can only be collected from features preserved in the fossil record (for example, in vertebrates, the hardest parts of the skeleton are more often preserved than soft parts; \citealt{sansomfossilization2013}). Therefore Total Evidence matrices are likely to contain a lot of missing data that may affect the method's ability to infer correct topologies, branch lengths and support values \citep{salamin2003}.

@@

Another practical implication of our results regards the tree inference methods. Because the Bayesian consensus trees consistently recovered topologies closer to the "best" tree topology than the Maximum Likelihood trees, we advise using Bayesian consensus trees to fix the topology where necessary for further steps in tree inference, for example in tip-dating \citep{ronquista2012,BEASTmaster}. Note that, although the Bayesian posterior tree distribution performed poorly in recovering the "best" tree topology, this is due to difficulties comparing distributions of trees (see discussion above). Thus, we do not suggest discarding the Bayesian posterior tree distribution once the topology has been fixed, particularly because these trees will be invaluable for phylogenetic comparative analyses \citep[e.g.][]{jetzthe2012}.

@@

Missing data in Total Evidence matrices is not a problem for recovering the "best" tree topology as long as enough living and fossil taxa in the matrix have data for overlapping morphological characters. When missing data increases in any of our missing data parameters ($M_{L}$, $M_{F}$ or $M_{C}$), it reduces support for the placement of fossil taxa and increases the displacement of wildcard taxa. Therefore we advise filling as many gaps in Total Evidence matrices as possible. Because this is difficult, if not impossible, for fossil data, we recommend coding as many morphological characters, for as many living taxa, as possible. Additionally, the topology of the Bayesian consensus trees, regardless the amount of missing data, were always closer to the "best" tree topology than the Maximum Likelihood trees. Therefore, we advise using Bayesian consensus tree topologies if topology must be fixed in a Total Evidence analysis, for example when using the tip-dating method. The results of our analyses are encouraging and show that it is possible to combine both neontological and palaeontological data in the same phylogeny despite issues of missing data. Hopefully, using these approaches will greatly improve our understanding of macroevolutionary processes through both time and space.
